---
title: "2 tipsport XML Feed"
author: "Ing. Peter Tomko, M.A."
date: "27/8/2020"
output:
  pdf_document: default
  html_document: default
---

```{r library, include = T, cache = F, comment = F, error = F, message = F, echo = T}
library(rvest)
library(stringr)
library(dplyr)
library(data.table)
library(xml2)
library(tidyverse)
library(DBI)
library(lubridate)

# - initialize the connection
con <- DBI::dbConnect(odbc::odbc(), "betting_ds")
```
Note: MIGHT BE REFRESHED ANYTIME.

# Introduction

The document provides an automated way of refreshing the data necessary for the project. There are two types of data source - historical match data `www.football-data.co.uk/data.php` and actual odds provided by `Tipsport`. 

## Tisport Data - Actual Bookmakers Odds

The betting company `Tipsport` provides the XML feed, in which the current bookmakers odds are included. The data source helps to create optimal investment strategy for the upcoming matches.

### Download and write into `t_tipsport_bookmaker`

Most of the obtained columns are described in the table creation script.

```{r tipsport data, warning = F, message = F, error = T, comment = F, prompt = F}
time_download <- Sys.time()

xml2_data <- xml2::download_xml("http://ban.tipsport.cz/f/oddsFeed.xml", 
                                file = "oddsFeed.xml")
tipsport_data <- xml2::read_html(xml2_data, encoding = "UTF-8")

df_supersport <- xml_find_all(tipsport_data, "//supersport") %>% 
  map_dfr(~ {
    # extract the attributes from the parent tag as a data.frame
    parent <- xml_attrs(.x) %>% enframe() %>% spread(name, value)
    
    # make a data.frame out of the attributes of the kids
    kids <- xml_children(.x) %>% map_dfr(~ as.list(xml_attrs(.x)))
    
    # combine them (bind_cols does not repeat parent rows)
    cbind.data.frame(parent, kids) %>% set_tidy_names() %>% as_tibble() }) %>%
  as.data.frame() %>%
  rename(sport = `name..1`,
         sport_category = `name..2`)

df_sport <- xml_find_all(tipsport_data, "//sport") %>% 
  map_dfr(~ {
    # extract the attributes from the parent tag as a data.frame
    parent <- xml_attrs(.x) %>% enframe() %>% spread(name, value)
    
    # make a data.frame out of the attributes of the kids
    kids <- xml_children(.x) %>% map_dfr(~ as.list(xml_attrs(.x)))
    
    # combine them (bind_cols does not repeat parent rows)
    cbind.data.frame(parent, kids) %>% set_tidy_names() %>% as_tibble() }) %>%
  as.data.frame() %>%
  rename(sport_category = `name..1`,
         sport_league = `name..3`,
         sport_id = `id`,
         sport_urlold = urlold,
         sport_url = url)

df_competition <- 
  data.frame(xml_find_all(tipsport_data, "//competition") %>% 
               map_dfr(~ {
                 # extract the attributes from the parent tag as a data.frame
                 parent <- 
                   xml_attrs(.x) %>% 
                   enframe() %>% 
                   spread(name, value)
                 
                 # make a data.frame out of the attributes of the kids
                 kids <- 
                   xml_children(.x) %>% 
                   map_dfr(~ as.list(xml_attrs(.x)))
                 
                 # combine them (bind_cols does not repeat parent rows)
                 cbind.data.frame(parent, kids) %>% 
                   set_tidy_names() %>% 
                   as_tibble() })) %>%
  
  as.data.frame() %>%
  rowwise() %>%
  mutate(urlold = na.omit(c(`urlold..11`, urlold..12)),
         url = na.omit(c(`url..12`, url..13)),
         sport_league = `name..3`,
         sport_url = `url..4`,
         sport_urlold = `urlold..5`,
         sport_id = `id..2`,
         competition_name = na.omit(c(`name..8`, `name..9`))) %>%
  select(-`id..6`, -`urlold..11`, -`url..12`, -`url..13`, -`name..3`, 
         -`url..4`, -`urlold..5`, -`id..2`, -`urlold..12`, -`name..8`,
         -`name..9`)

df_match <- xml_find_all(tipsport_data, "//match") %>% 
  map_dfr(~ {
    # extract the attributes from the parent tag as a data.frame
    parent <- xml_attrs(.x) %>% enframe() %>% spread(name, value)
    
    # make a data.frame out of the attributes of the kids
    kids <- xml_children(.x) %>% map_dfr(~ as.list(xml_attrs(.x)))
    
    # combine them (bind_cols does not repeat parent rows)
    cbind.data.frame(parent, kids) %>% set_tidy_names() %>% as_tibble() }) %>%
  as.data.frame() %>%
  rowwise() %>%
  mutate(eventid = na.omit(c(`id..8`, `id..9`)),
         eventname = na.omit(c(`name..9`, `name..10`)),
         competition_name = na.omit(c(`name..4`, `name..5`))) %>%
  select(-`id..2`, -`id..8`, -`id..9`, -`name..9`, -`name..10`, 
         -`name..4`, -`name..5`)

df_event <- xml_find_all(tipsport_data, "//event") %>% 
  map_dfr(~ {
    # extract the attributes from the parent tag as a data.frame
    parent <- xml_attrs(.x) %>% enframe() %>% spread(name, value)
    
    # make a data.frame out of the attributes of the kids
    kids <- xml_children(.x) %>% map_dfr(~ as.list(xml_attrs(.x)))
    
    # combine them (bind_cols does not repeat parent rows)
    cbind.data.frame(parent, kids) %>% set_tidy_names() %>% as_tibble() }) %>%
  as.data.frame() %>%
  rename(eventid = `id..4`,
         eventname = name,
         opportunityid = `id..6`)

t_tipsport_bookmaker <- 
  df_event %>%
  left_join(., df_match) %>%
  left_join(., df_competition) %>%
  left_join(., df_sport) %>%
  left_join(., df_supersport) %>%
  mutate(rate = as.numeric(rate),
         created_at = time_download,
         dateclosed = dmy_hm(dateclosed)) %>%
  as.data.frame()

dbWriteTable(con, "t_tipsport_bookmaker", t_tipsport_bookmaker, 
             row.names = F, append = T)

rm(df_competition)
rm(df_event)
rm(df_match)
rm(df_sport)
rm(df_supersport)
rm(tipsport_data)
rm(time_download)
rm(xml2_data)
```
